---
title: "Working with factors: the forcats package"
date: "111022 | Data management and visualization with R"
author: 
  - "Aitor Ameztegui"
  - "Marcos Rodrigues"
output:
  rmdformats::readthedown:
    highlight: pygments
    number_sections: true
    css: lab.css
# runtime: shiny_prerendered
---

```{r include=FALSE}
library(tidyverse)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
#knitr::opts_chunk$set(eval = FALSE)
load('data/data_IFN.Rdata')
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
load('data/data_IFN.Rdata')
options(width = 85)
```

# Introduction. Create factors

In R, factors are  a special case of categorical variables, in which variables have a fixed and known set of possible values. Historically, factors were much easier to work with than characters. As a result, many of the functions in base R automatically convert characters to factors. This means that factors often crop up in places where they’re not actually helpful. Fortunately, you don’t need to worry about that in the tidyverse, and can focus on situations where factors are genuinely useful.

## Creating factors

Imagine that you have a variable that records month names:

```{r pines}

x1 <- c("Dec", "Apr", "Jan", "Mar")

```


Using a string to record this variable has two problems:

1. The character string doesn’t sort in a useful way:

```{r}
sort(x1)
```

2. There are only twelve possible months, but there’s nothing saving you from typos:

```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
```

Both problems can be fixed with a factor. To create a factor you must start by creating a list of the valid levels:

```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```

Now you can create the factor, indicating which levels will be allowed:

```{r}
y1 <- factor(x1, levels = month_levels)
y1
```

Now, if we sort the values:

```{r}
sort(y1)

```

And any values not in the set (or mistakes) will be silently converted to `NA:
```{r}
y2 <- factor(x2, levels = month_levels)
y2
```

If you omit the levels, they’ll be taken from the data in alphabetical order:

```{r}
factor(c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))

```

Sometimes you’d prefer that the order of the levels match the order of the first appearance in the data. You can do that when creating the factor by setting levels to `unique(x)`, or after the fact, with `fct_inorder()`:

```{r}
f1 <- factor(x1, levels = unique(x1))
f1

f2 <- x1 %>% factor() %>% fct_inorder()
f2
```

If you ever need to access the set of valid levels directly, you can do so with `levels()`:

```{r}
levels(f2)
```


# Modifying factor order

It’s often useful to change the order of the factor levels in a visualisation. For example, imagine you want to explore the mean height of several tree species

```{r}


mean_height <- trees %>%
    group_by(Especie) %>%
    summarise(height = mean(HeiIf3, na.rm = TRUE))

ggplot(mean_height, aes(height, Especie)) + geom_point() 
```

It is difficult to interpret this plot because there’s no overall pattern. We can improve it by reordering the levels of `Especie` using `fct_reorder()`. `fct_reorder()` takes three arguments:

- `f`, the factor whose levels you want to modify.
- `x`, a numeric vector that you want to use to reorder the levels.
- Optionally, `fun`, a function that’s used if there are multiple values of `x` for each value of `f`. The default value is median.

```{r}

mean_height <- mean_height %>%
    mutate(Especie = fct_reorder(Especie, height)) 

ggplot(mean_height, aes(height, Especie)) +
    geom_point()
```


What if we create a similar plot looking at how long it took to measure plots across reported income level?

```{r}
rincome_summary <- gss_cat %>%
  group_by(rincome) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(rincome_summary, aes(age, fct_reorder(rincome, age))) + geom_point()
```

Here, arbitrarily reordering the levels isn’t a good idea! That’s because rincome already has a principled order that we shouldn’t mess with. Reserve fct_reorder() for factors whose levels are arbitrarily ordered.

However, it does make sense to pull “Not applicable” to the front with the other special levels. You can use fct_relevel(). It takes a factor, f, and then any number of levels that you want to move to the front of the line.
```{r}
ggplot(rincome_summary, aes(age, fct_relevel(rincome, "Not applicable"))) +
  geom_point()
```

Another type of reordering is useful when you are colouring the lines on a plot. fct_reorder2() reorders the factor by the y values associated with the largest x values. This makes the plot easier to read because the line colours line up with the legend.

```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  count(age, marital) %>%
  group_by(age) %>%
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop, colour = marital)) +
  geom_line(na.rm = TRUE)

ggplot(by_age, aes(age, prop, colour = fct_reorder2(marital, age, prop))) +
  geom_line() +
  labs(colour = "marital")
```

Finally, for bar plots, you can use fct_infreq() to order levels in increasing frequency: this is the simplest type of reordering because it doesn’t need any extra variables. You may want to combine with fct_rev().

```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital)) +
    geom_bar()
```


# Modifying factor levels

More powerful than changing the orders of the levels is changing their values. This allows you to clarify labels for publication, and collapse levels for high-level displays. The most general and powerful tool is fct_recode(). It allows you to recode, or change, the value of each level. For example, take the gss_cat$partyid:

```{r}
gss_cat %>% count(partyid)

```
The levels are terse and inconsistent. Let’s tweak them to be longer and use a parallel construction.

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)

```

fct_recode() will leave levels that aren’t explicitly mentioned as is, and will warn you if you accidentally refer to a level that doesn’t exist.

To combine groups, you can assign multiple old levels to the same new level:

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
  )) %>%
  count(partyid)
```

You must use this technique with care: if you group together categories that are truly different you will end up with misleading results.

If you want to collapse a lot of levels, fct_collapse() is a useful variant of fct_recode(). For each new variable, you can provide a vector of old levels:

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

Sometimes you just want to lump together all the small groups to make a plot or table simpler. That’s the job of fct_lump():


```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)
```

The default behaviour is to progressively lump together the smallest groups, ensuring that the aggregate is still the smallest group. In this case it’s not very helpful: it is true that the majority of Americans in this survey are Protestant, but we’ve probably over collapsed.

Instead, we can use the n parameter to specify how many groups (excluding other) we want to keep:

```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 10)) %>%
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

