---
title: "Lab 7.1 - Visualizing spatial information"
date: "111022 | Data management and visualization with R"
author: 
  - "Aitor Ameztegui"
bibliography: "references/biblio.bib"
output:
  rmdformats::readthedown:
    highlight: pygments
    number_sections: true
    css: lab.css
# runtime: shiny_prerendered
---


```{r knitr_init, echo=FALSE, cache=FALSE, warning=FALSE}

library(knitr)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


```{r include=FALSE}
library(tidyverse)
library(sf)
library(raster)

options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE     # for regression output
  )
fires <- read.csv('data/fires.csv')
grid <- read_sf('data/Fires_grid_CCAA.shp')
```

# Accessing spatial data

In addition to the functions for statistical calculation that we have already seen, in R we can find several packages that allow to interact with spatial data, both in vector and raster format. Some of the main GIS packages in R are:

- `rgdal`
- `Maptools`
- `raster`
- `GISTools`
- `Sp`
- `sf`

Basically, these are libraries (packages) allow as reading and managing spatial data objects, such as ESRI shapefiles or raster files in ASCII format or any other GDAL supported format. These packages also allow you to create thematic maps from spatial data or to perform most of the geoprocesses of a Geographic Information System (GIS).

## Working with vector layers

The `sf` package will allow us, among other things, to access information contained in spatial vector layers (points, lines and polygons). This package incorporates many other functions for spatial data manipulation, but they are way beyond the scope of this course.

Vector information is characterized by representing elements by points, lines and polygons. Each of these elements is related to a record within an attribute table associated with spatial information. If you are not familiar with GIS or spatial data take a quick look to http://gisgeography.com/spatial-data-types-vector-raster/.

By reading vectors into R we can access both the spatial information (location) and thematic (attribute table). But the coolest thing about `sf` it's not that translate vector layers into `data.frames`. The best thing above all is that **it is fully compatible with `dplyr`**. Contrary to other packages like `rgdal`, `sf` implements a class object that allows using pipes and most functions in `dplyr` using variables from the vector attribute table.

To read a vector `shape` layer into R we proceed as follows:

```{r}
grid <- read_sf('data/Fires_grid_CCAA.shp')
```

The previous chunk of code loads the layer *Fires_grid_CCAA* into an R object we called `grid`. This layer contains a grid polygon division that matches the `CODCUAD` identifier in the `fires` data frame.

> Note that shapefiles are composed by many individual files, sharing the name but with differnt file extensions

Next, we will see how to use `ggplot` for mapping. It is actually similar to what we have done so far. The only specific thing is the use of `geom_sf' to display spatial vector layers. You will notice how once we are within that particular geom we can use regular aesthetics to display info.

```{r}
#First we summarise number of fires and burned area by CODCUAD
fire.features.grid <- fires %>%
  filter(YEAR>1994,BAREA>1) %>%
  group_by(CODCUAD,.drop = F) %>%
  summarise(N = n(),BA = sum(BAREA))

#Then we join the fire dataset to the grid
grid %>%
    left_join(fire.features.grid,by=c('Cod_Cuad'='CODCUAD')) %>%
    ggplot() +
        geom_sf(aes(fill=log(BA)))
    
```

Notice how the results slightly differs from what we saw in regular plots. The axis labels are now geographic coordinates rather than unit coordinates. Other than that, the plot behaves the same.

```{r}
#First we summarise number of fires and burned area by CODCUAD
fire.features.grid <- fires %>%
  filter(YEAR>1994,BAREA>1) %>%
  group_by(CODCUAD,.drop = F) %>%
  summarise(N = n(),BA = sum(BAREA))

#Then we join the fire dataset to the grid
grid %>%
    left_join(fire.features.grid,by=c('Cod_Cuad'='CODCUAD')) %>%
    ggplot() +
        geom_sf(aes(fill=log(BA)), color =NA)
    
```

## Working with raster layers

We have already seen how to access spatial information (GIS) in vector format using the `maptools` package. With this we could already read information from a layer in shapefile format and include it in a regression script as input data to create the dependent variable. In the next section we'll do that.

The only thing we would have to do is to introduce the information related to the independent variables. For this we will use some of the functionalities available in the raster package. In this way we will also automate the obtaining of the independent variables.

The `raster` package provides utilities for manipulating geographic information in raster format. This package, along with maptools, converts R into a pseudo-GIS environment, allowing you to use most basic raster analysis functions. The raster package allows:

- Create raster objects.
- Manipulation of very heavy files.
- Map algebra and overlay functions.
- Distances and connection analysis.
- Conversion of polygons, lines and points to raster.
- Calculate __spatial predictions__ of regression models.
- Mapping from raster maps.
- Reading and writing various types of raster files.

Of all these functionalities available we will focus on:

- Creating raster objects.
- Calculate predictions of regression models.
- Elaboration of maps.
- Reading and writing various types of raster files.

The remainnig functions can be found in a more intuitive way in any desktop GIS software. However, in some cases it may be more operative to use them directly in R.

The information in raster format is basically a 2-dimensional matrix, in which each of the grids or pixels is given a numerical value that represents either a category o parameter. This type of structure is particularly suitable to represent information about phenomena that has value in any location (continuous) such as elevation, temperature, distances... If we recall the independent variables used in the regression practice, they referred to phenomena of this nature. What we will do next is using the *shapefile* layer of weather stations to get information regarding the independent variables from raster layers. 

### Loading raster layers

Again, the most first thing to do is reading a raster layer. Once `raster` package is installed and loade we can call the `raster()` function and load any raster layer. Let's see an example using the layer *dem_rioja.tif* that you can find in the folder *Raster*. It is a Digital Elevation Model (DEM)^[A DEM is a raster layer that takes elevation value in each pixel.] of the Autonomous Community of La Rioja^[retrieved from http://centrodedescargas.cnig.es/CentroDescargas/index.jsp].

The first thing is to install the package. Unlike in other occasions this time we will specify the argument `dep = TRUE`. The raster package has many dependencies, ie, needs functions from other packages in order to work properly. It could be very annoying installing them one by one so we use that extra argument to ensure everything will work fine:

```{r eval=FALSE}
# install.packages('raster', dep = TRUE)
library(raster)
```

Once installed and activated we can call the `raster()` function to read raster layer. 

```{r}
dem <- raster('data/dem_rioja.tif')
```

Once we loaded the raster layer we are ready to display it using ggplot. There are two ways of doing so. The *canonic* way involves converting the raster layer into a regular data frame and then use `geom_raster` or `geom_tile`. A raster layer is essentially a matrix where each pixel can be individualized using coordinates and the value that the pixel holds. That produces a dataframe with x, y variables and a third z variable with the variable contained in the raster:

```{r}
raster::as.data.frame(dem, xy = TRUE) %>%
    ggplot() +
        geom_raster(aes(x=x,y=y,fill=dem_rioja))

```

The second option is using the function `gplot` from package `rasterVis`. This allows using raster objects as data in the plot call. Other than that, the `gplot` function works the same `ggplot` does.

```{r}
library(rasterVis)
gplot(dem) +
    geom_tile(aes(fill=value)) 
```

# The elements of a map

- Legend: symbology key to understand the elements and data in a map (or plot)
- Scale: representation of the unit ratio map-reality
- North: indication of the geographical norht position to understand the arrangment of the map.

# Spatial properties and characteristics

To end the basics of access to spatial layers let's focus on seeing some fundamental properties that share all spatial layers. In essence we will see how to obtain and modify some parameters such as the extension of the layers or the reference system. In future modules we will go deeper into these concepts.


## Layer extent

The layer extent is the rectangle that sorrounds spatial features, ie, the limits of data region. We can access the spatial extent of any layers with the `extent()` function. This function, which is common for both vector and raster layers, returns the coordinates of the ends of the layers:

```{r}
extent(dem)
extent(grid)
```

What good is this for us? Well, this can be used to know if our layers spatially match, to save this information and use it to cut layers or create empty raster "canvases" on which to dump information, for example the result of an interpolation process. In the short term, the greater utility of this function is to find out the units of the reference system in case we do not know the information.

## Projections and reference system

The reference system (CRS) is a key element of spatial information. So far we have worked the spatial information without paying attention to this parameter. However, sooner or later we'll have to deal with it. Not having taken this into account will give us problems. For example, we will not be able to overlay or combine layers by the absence or difference of reference systems, so we need to know how assing or reproject a layer to a different reference system.

### Assign a Coordinate Reference System

A Coordinate Reference System or CRS must only be assign when a spatial layer lacks this information. Therefore, first  we will see if our layers have or have not been assigned a reference system. A single function to do allows check and assing the CRS, `crs()`. We will apply this function on the layers `wStations` and `dem`:

```{r}
crs(dem)
crs(grid)
```

In this particular case none of the layers were assigned a reference system. To give one first we have to know what system corresponds to them and how to assign it. The system that corresponds to them is the EPSG: 23030 - UTM ED50 30N. You can get a list of the different ways of coding reference systems at http://spatialreference.org/. In this case we will use the parameters according to the coding proj4 which is:

```{r eval=FALSE}
+proj=utm +zone=30 +ellps=intl +units=m +no_defs

```

To assign the CRS we proceed as follows:

```{r}
crs(dem)<-"+proj=utm +zone=30 +ellps=intl +units=m +no_defs"

```

Finally, we check the CRS again:

```{r}
crs(dem)

```

### Layer projection

But, what if we want to change the CRS? Obviously in this case we can not use the above procedure. Furthermore, in this case, the same procedure is not used for each data model (vector or raster), but there is a function for the case of vector layers and another for raster layers. In the case of the vector layers we proceed as follows:

```{r}
grid.ETRS89<-st_transform(grid,"+proj=utm +zone=30 +ellps=GRS80 +units=m +no_defs ")
crs(grid.ETRS89)

```

For raster layers the procedure is similar but the funcion changes:

```{r}
dem.ETRS89<-projectRaster(dem,crs=crs(grid.ETRS89))
crs(dem.ETRS89)

```
Note that in the raster example we retrieved the CRS from an existing layer rather than specify the CRS raw. 

Something that is worth mention now is that we cannot project a layer that lacks CRS, first we have to assign it one.

<div class= "exercise">
**EXERCISE 1**</br>

Map the following fire regime features using data from `fires.csv`:

- Proportion of fires occurring during summer (June to September)
- Proportion of natural fires
- 95th percentile of fire size in each CCAA

</div>


```{r}
ex1_a <- fires %>% 
  group_by(CODCUAD) %>% 
  summarise(N=n(), summer_fire = n_distinct(IDPIF[between(MONTH, 6, 9)]),
            prop = (summer_fire/N)*100)

grid %>%
  left_join(ex1_a, by = c(Cod_Cuad = "CODCUAD")) %>%
  ggplot() + geom_sf(aes(fill = prop))
```

```{r}
# proportion of natural fires

ex1_b <- fires %>% 
  group_by(CODCUAD) %>% 
  summarise(N=n(), nat_fire = n_distinct(IDPIF[CAUSE == 1]),
            prop = (nat_fire/N)*100)

grid %>%
  left_join(ex1_b, by = c(Cod_Cuad = "CODCUAD")) %>%
  ggplot() + geom_sf(aes(fill = prop))

```

```{r}
# 95th percentile of fire size in each CCAA

fires_p95 <- fires %>%  
  group_by(CCAA) %>%
  summarise(quant95 = quantile(BAREA, 0.95))

ccaa_fires <- fires %>%
  left_join(fires_p95) 

fires95_grid <- ccaa_fires %>% 
  group_by(CODCUAD, CCAA) %>%
  summarise(quant95 = dplyr::first(quant95))

# Then we join the fire dataset to the grid
grid %>%
  left_join(fires95_grid, by = c(Cod_Cuad = "CODCUAD")) %>%
  ggplot() + geom_sf(aes(fill = quant95))


```



