---
title: "Welcome to </br>Data Management </br>and Visualization with R"
author: "Dr. Aitor Ameztegui"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/udl_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

# Who are we?


.pull-left[.bold[
 [Dr. Aitor Ameztegui](http://ameztegui.weebly.com/)<br>]

<i class="fa fa-envelope"></i> &nbsp; [aitor.ameztegui@eagrof.udl.cat](mailto:aitor.ameztegui@eagrof.udl.cat) <br>
<i class="fa fa-university"></i> &nbsp; Building 4, Bureau 2.04.1 <br>
<i class="fa fa-phone"></i> &nbsp; 973702540 <br> 
<i class="fa fa-calendar"></i> &nbsp; Tue 11:00 - 12:30 <br> 
&nbsp; &nbsp; &nbsp; Thur 15:00 - 16:00 or <br>
&nbsp; &nbsp; &nbsp; upon request
<br><br>]

.pull-right[.bold[
[Dr. Marcos Rodrigues](http://www.eagrof.udl.cat/es/personal/PDI/rodrigues.html) <br>]

<i class="fa fa-envelope"></i> &nbsp; [rmarcos@eagrof.udl.cat](mailto:rmarcos@eagrof.udl.cat) <br>
<i class="fa fa-university"></i> &nbsp; Building 4, Bureau 2.05.1 <br>
<i class="fa fa-phone"></i> &nbsp; 973702546 <br> 
<i class="fa fa-calendar"></i> &nbsp; Mon 12:00 - 2:00 or <br>
&nbsp; &nbsp; &nbsp; upon request
<br><br>]

---


# What this course is about?

- Everything you want to know about the course, and everything you will need for the course will be posted at [datamanagement.netlify.com/](https://datamanagement.netlify.com/)

--

This course is about: 

- <i class="fa fa-code fa-10px"></i> Coding

--

- <i class="fa fa-database fa-10px"></i> Data management

--

- <i class="fa fa-bar-chart fa-10px"></i> Data visualization

--

- <i class="fa fa-flask fa-10px"></i> Modelling and statistics

--

- <i class="fa fa-users fa-10px"></i> Interpretation

--

-   <i class="fa fa-code fa-10px"></i> + 
    <i class="fa fa-database fa-10px"></i> +
    <i class="fa fa-bar-chart fa-10px"></i> +
    <i class="fa fa-flask fa-10px"></i> +
    <i class="fa fa-users fa-10px"></i> +
    = Data Science!
--

<br>

**Data science** is an exciting discipline that allows you to turn raw data into understanding, insight, and knowledge. We're going to learn to do this in a `tidy` way -- more on that later!


---

# What this course is NOT about?


- R

--

- Statistics

--

- Modelling

--

- Big data


---

# Why use R?

--

- What's R?

--

- R is incredibly versatile

--

- R is a true “data language"

--

- R is multiplatform

--

- R is 100% free (in both senses)

--

- Analyses conducted in R are transparent, easily shareable, and reproducible

--

- R has a huge support community

---
class: center, middle, inverse


# Let'see some examples of data in the wild


---
# Example 1
## A year as told by fitbit

### by Nick Strayer


http://livefreeordichotomize.com/2017/12/27/a-year-as-told-by-fitbit/

---
# Example 2

## R-Ladies global tour

### by Maelle Salmon

http://www.masalmon.eu/2017/10/06/globalrladiestour/

---

# Example 3

## Text analysis of Trump's tweets confirms he writes only the (angrier) Android half

### by David Robinson (Stack Overflow)

http://varianceexplained.org/r/trump-tweets/

---

class: center, middle, inverse

# Your turn!


---

# Join RStudio.cloud

- What's RStudio?

--

- And why RStudio Cloud?

--

Go to [shorturl.at/gitGM](shorturl.at/gitGM), and log in.

<font color="#E34132">
Once done, place a green sticky on your laptop. If you have questions, place a pink sticky.
</font>
---

# Create your first data visualization

- Once you log on to RStudio Cloud, click on this course's workspace "111022 Data Management and Visualization"

- You should see a project called `SIBECOL twitter analysis`

- In the Files pane in the bottom right corner, spot the file called `index.Rmd`. Open it, and then click on the "Knit" button.

- Go back to the file and change your name on top (in the `yaml` -- we'll talk about what this means later) and knit again.

- Then change the country names to those you're interested in. Your spelling and capitalization should match the data so take a peek at the Appendix to see how the country names are spelled. Knit again. And voila, your first data visualization!

<font color="#E34132">
Once done, place a green sticky on your laptop. If you have questions, place a pink sticky.
</font>
---

class: center, middle, inverse

# Course structure and policies

---

# Class meetings

- Interactive

- There will be some lectures, lots of learn-by-doing

- Bring your laptop to class every day


---

---

# Course components:

- Lectures: we will try to keep them short, but some "theory" will be needed

- Labs: hands-on application of concepts and tools seen in class. 
    - They're self paced tutorials with examples
    - There will be some questions/exercises you will need to answer, by creating a "lab report" (we'll see how). Try to finish in class, but can continue later. Check/no check
- Homework: Individual. You are expected to work at home (but can ask doubts of course!)
- Exams: Individual, two take-home midterms
- Final project: during the course eahc of you will design a data science project using your wn dataset (ideally for your master's thesis). You must present the results to your classmates. You must complete the final project and be in class to present it to pass this class.

---

# Grading

Component             | Weight
----------------------|----------------
Participation & application exercises                   | 10%*
Peer evaluation       | 5%
Homework              | 20%
Labs                  | 15%
Midterm 1             | 17.5%
Midterm 2             | 17.5%
Final project         | 15%

</br>
- Class attendance is a firm expectation; frequent absences or tardiness 
will be considered a legitimate cause for grade reduction.

- Excused absences do not excuse you from assigned homework, it is your responsibility to make alternative arrangements to turn in any assignments in a timely fashion.

---
# Late/missed work policy

- Late work policy for homework assignments:
    - late, but within 48 hours of due date/time: -20%
    - any later: no credit

- If you cannot complete an assignment on the due date due to a short-term 
illness, tell us in advance and we can arrange some extra time, then the regular late work policy kicks in.

- Late work will not be accepted for take home midterms and the final project.

- If you are faced with a personal or family emergency or a long-range or 
chronic health condition that interferes with your ability to attend or 
complete classes, please talk to us.

- You must ABSOLUTELY complete the final project and be in class to present it in order to pass this course.

---

# Academic integrity

> To uphold the Duke Community Standard:

> - I will not lie, cheat, or steal in my academic endeavors;
> - I will conduct myself honorably in all my endeavors; and
> - I will act if the Standard is compromised.

- Only work that is clearly assigned as team work can be completed collaboratively.

- Use of disallowed materials during the take home exam will not be tolerated.

---

# Sharing/reusing code

- We are aware that a huge volume of code is available on the web to solve any number of problems.

- Unless we explicitly tell you not to use something, the course's policy is that you may make use of any online resources (e.g. StackOverflow) but you must explicitly cite where you obtained any code you directly use (or use as inspiration). 

- Any recycled code that is discovered and is not explicitly cited will be treated as plagiarism. 
- Except for the take home exams, you are welcome to discuss the problems together and ask for advice, but you may not directly send or make use of code from another student in this class 

- On the take home exams all communication with classmates is explicitly forbidden.

---

# Diversity & Inclusiveness:

- Intent: Students from all diverse backgrounds and perspectives be well-served by this course, that students' learning needs be addressed both in and out of class, and that the diversity that the students bring to this class be viewed as a resource, strength and benefit. It is my intent to present materials and activities that are respectful of diversity: gender identity, sexuality, disability, age, socioeconomic status, ethnicity, race, nationality, religion, and culture. Let me know ways to improve the effectiveness of the course for you personally, or for other students or student groups.

- If you have a name and/or set of pronouns that differ from those that appear in your official UdL records, please let us know!

- If you feel like your performance in the class is being impacted by your experiences outside of class, please don't hesitate to come and talk with any of us If you prefer to speak with someone outside of the course, the MSc coordinator is an excellent resource. 

- We (like many people) are still in the process of learning about diverse perspectives and identities. If something was said in class (by anyone) that made you feel uncomfortable, please talk to us about it.

---

# How to get help

MANY RESOURCES!!!

- Course content, logistics, etc. discussion via GitHub on the [Sta199-S18/community](https://github.com/Sta199-S18/community) repository. Note that this is a public discussion forum, which means others outside of the course can stumble upon it and help you as well.
    - See course syllabus for tips on posting questions.

- For personal and grade related questions, use email.

---
class: center, middle, inverse

# Questions?
