---
title: "Good coding practices"
author: "Dr. Aitor Ameztegui"
date: "2019-09-15"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/sta199-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1)
opts_chunk$set(eval = FALSE)

# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```
---

# Good code vs. bad code

* There is no such thing as inherently bad code

* However, this does not mean we shouldn't judge our code, and try to improve our coding practices

* Some decisions genuinely do make code easier to use (especially matching indenting to programming structure), but many decisions are arbitrary

* Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread

---
 # Good code vs. bad code

* We're gonna cover some good coding practices
* This includes style
* But also bla bla bla
* Some f the things you will not understand now, but it is important to see them before
* It will act as a repository
* Coding style recommendations will be fundamentaly based on the tidyverse style guide, by Wickham (https://style.tidyverse.org/)


---

# Good coding style: Files

## File names

File names should be meaningful and end in `.R`. Avoid using special characters in file names - stick with numbers, letters, `-`, and `_`.

.pull-left[
```{r, eval=FALSE}
# Good
fit_models.R
utility_functions.R
```
]

.pull-right[

```{r, eval=FALSE}
# Bad
fit models.R
foo.r
stuff.r
```
]

---

# Good coding style: Files

## File names

If files should be run in a particular order, prefix them with numbers. If it seems likely you’ll have more than 10 files, left pad with zero:

```{r, eval=FALSE}
00_download.R
01_explore.R
...
09_model.R
10_visualize.R
```

Pay attention to capitalization. Prefer file names that are all lower case, and never have names that differ only in their capitalization.

---
# Good coding style: Files

## RStudio set up

---

# Good coding style: Files

## Internal structure

Use commented lines of `-` to break up your file into easily readable chunks.

```{r, eval= FALSE}

# Load data ---------------------------

some code
some more code

# Plot data ---------------------------

still some more code
```

If your script uses add-on packages, load them all at once at the very beginning of the file. This is more transparent than sprinkling `library()` calls throughout your code

---

# Good coding style: Objects

## Object names

Variable and function names should use only lowercase letters, numbers, and `_`. Use underscores (`_`) (so called snake case) to separate words within a name.

```{r}
# Good
day_one
day_1

# Bad
DayOne
dayone
```

Base R uses dots in function names (`contrib.url()`) and class names (`data.frame`), but it’s better to reserve dots exclusively for the S3 object system. In S3, methods are given the name `function.class`; if you also use `.` in function and class names, you end up with confusing methods like `as.data.frame.data.frame()`.

If you find yourself attempting to cram data into variable names (e.g. `model_2018`, `model_2019`, `model_2020`), consider using a list or data frame instead.

Generally, variable names should be nouns and function names should be verbs. Strive for names that are concise and meaningful (this is not easy!).

```{r}
# Good
day_one

# Bad
first_day_of_the_month
djm1
```

Where possible, avoid re-using names of common functions and variables. This will cause confusion for the readers of your code.
```{r}
# Bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```



---
# Good coding style: Objects

## Spacing

### Commas

Always put a space after a comma, never before, just like in regular English.

```{r}
# Good
x[, 1]

# Bad
x[,1]
x[ ,1]
x[ , 1]

```

---

# Good coding style: Objects

## Spacing

### Parentheses

Do not put spaces inside or outside parentheses for regular function calls.

```{r}
# Good
mean(x, na.rm = TRUE)

# Bad
mean (x, na.rm = TRUE)
mean( x, na.rm = TRUE )
```


Place a space before and after () when used with if, for, or while.

```{r}
# Good
if (debug) {
  show(x)
}

# Bad
if(debug){
  show(x)
}
```


Place a space after () used for function arguments:

```{r}
# Good
function(x) {}

# Bad
function (x) {}
function(x){}
```

---

# Good coding style: Objects

## Spacing

### Infix operators

Most infix operators (`==`, `+`, `-`, `<-`, etc.) should always be surrounded by spaces:

```{r}
# Good
height <- (feet * 12) + inches
mean(x, na.rm = 10)

# Bad
height<-feet*12+inches
mean(x, na.rm=10)
```

---

# Good coding style: Objects

## Spacing

### Infix operators

Adding extra spaces ok if it improves alignment of `= `or `<-`.

```{r}
# Good
list(
  total = a + b + c,
  mean  = (a + b + c) / n
)

# Also fine
list(
  total = a + b + c,
  mean = (a + b + c) / n
)
```

Do not add extra spaces to places where space is not usually allowed.

---
# Good coding style: Objects

## Code blocks

Curly braces, `{}`, define the most important hierarchy of R code. To make this hierarchy easy to see:

`{` should be the last character on the line. Related code (e.g., an if clause, a function declaration, a trailing comma, …) must be on the same line as the opening brace.

The contents should be indented by two spaces.

`}` should be the first character on the line.

```{r}

# Good
if (y < 0 && debug) {
  message("y is negative")
}

if (y == 0) {
  if (x > 0) {
    log(x)
  } else {
    message("x is negative or zero")
  }
} else {
  y^x
}

test_that("call1 returns an ordered factor", {
  expect_s3_class(call1(x, y), c("factor", "ordered"))
})

tryCatch(
  {
    x <- scan()
    cat("Total: ", sum(x), "\n", sep = "")
  },
  interrupt = function(e) {
    message("Aborted by user")
  }
)

# Bad
if (y < 0 && debug) {
message("Y is negative")
}

if (y == 0)
{
    if (x > 0) {
      log(x)
    } else {
  message("x is negative or zero")
    }
} else { y ^ x }

```

---

## Inline statements

It’s ok to drop the curly braces for very simple statements that fit on one line, as long as they don’t have side-effects.

```{r}

# Good
y <- 10
x <- if (y < 20) "Too low" else "Too high"
Function calls that affect control flow (like return(), stop() or continue) should always go in their own {} block:

# Good
if (y < 0) {
  stop("Y is negative")
}

find_abs <- function(x) {
  if (x > 0) {
    return(x)
  }
  x * -1
}

# Bad
if (y < 0) stop("Y is negative")

if (y < 0)
  stop("Y is negative")

find_abs <- function(x) {
  if (x > 0) return(x)
  x * -1
}

```


---

## Lines

Strive to limit your code to 80 characters per line. This fits comfortably on a printed page with a reasonably sized font. If you find yourself running out of room, this is a good indication that you should encapsulate some of the work in a separate function.

If a function call is too long to fit on a single line, use one line each for the function name, each argument, and the closing `)`. This makes the code easier to read and to change later.

```{r}
# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
                              )
```

As described under Argument names, you can omit the argument names for very common arguments (i.e. for arguments that are used in almost every invocation of the function). Short unnamed arguments can also go on the same line as the function name, even if the whole function call spans multiple lines.

```{r}
map(x, f,
  extra_argument_a = 10,
  extra_argument_b = c(1, 43, 390, 210209)
)
```


You may also place several arguments on the same line if they are closely related to each other, e.g., strings in calls to `paste()` or `stop()`. When building strings, where possible match one line of code to one line of output.

```{r}
# Good
paste0(
  "Requirement: ", requires, "\n",
  "Result: ", result, "\n"
)

# Bad
paste0(
  "Requirement: ", requires,
  "\n", "Result: ",
  result, "\n")
```


---

## Assignments

Use `<-`, not `=`, for assignment.

```{r}
# Good
x <- 5

# Bad
x = 5
```



---
  
## Comments

In data analysis code, use comments to record important findings and analysis decisions. If you need comments to explain what your code is doing, consider rewriting your code to be clearer. If you discover that you have more comments than code, considering switching to RMarkdown.


---

# Functions

## Naming

As well as following the general advice for object names, strive to use verbs for function names:

```{r}
# Good
add_row()
permute()

# Bad
row_adder()
permutation()

```

---

## Long lines

If a function definition runs over multiple lines, indent the second line to where the definition starts.

```{r}

# Good
long_function_name <- function(a = "a long argument",
                               b = "another argument",
                               c = "another long argument") {
  # As usual code is indented by two spaces.
}

# Bad
long_function_name <- function(a = "a long argument",
  b = "another argument",
  c = "another long argument") {
  # Here it's hard to spot where the definition ends and the
  # code begins
}
```

---

# Pipes

## Introduction
Use `%>%` to emphasise a sequence of actions, rather than the object that the actions are being performed on.

Avoid using the pipe when:

* You need to manipulate more than one object at a time. Reserve pipes for a sequence of steps applied to one primary object.

* There are meaningful intermediate objects that could be given informative names.
---

# Pipes

## Whitespace
`%>%` should always have a space before it, and should usually be followed by a new line. After the first step, each line should be indented by two spaces. This structure makes it easier to add new steps (or rearrange existing steps) and harder to overlook a step.

```{r}
# Good
iris %>%
  group_by(Species) %>%
  summarize_if(is.numeric, mean) %>%
  ungroup() %>%
  gather(measure, value, -Species) %>%
  arrange(value)

# Bad
iris %>% group_by(Species) %>% summarize_all(mean) %>%
ungroup %>% gather(measure, value, -Species) %>%
arrange(value)
```

---
# Pipes
## Long lines
If the arguments to a function don’t all fit on one line, put each argument on its own line and indent:

```{r}
iris %>%
  group_by(Species) %>%
  summarise(
    Sepal.Length = mean(Sepal.Length),
    Sepal.Width = mean(Sepal.Width),
    Species = n_distinct(Species)
  )
```



---
# Pipes
## Short pipes

A one-step pipe can stay on one line, but unless you plan to expand it later on, you should consider rewriting it to a regular function call.

```{r}
# Good
iris %>% arrange(Species)

iris %>% 
  arrange(Species)

arrange(iris, Species)
```


Sometimes it’s useful to include a short pipe as an argument to a function in a longer pipe. Carefully consider whether the code is more readable with a short inline pipe (which doesn’t require a lookup elsewhere) or if it’s better to move the code outside the pipe and give it an evocative name.

```{r}

# Good
x %>%
  select(a, b, w) %>%
  left_join(y %>% select(a, b, v), by = c("a", "b"))

# Better
x_join <- x %>% select(a, b, w)
y_join <- y %>% select(a, b, v)
left_join(x_join, y_join, by = c("a", "b"))

```

---

## Assignments

There are three acceptable forms of assignment:

1. Variable name and assignment on separate lines:

```{r}
iris_long <-
  iris %>%
  gather(measure, value, -Species) %>%
  arrange(-value)
```

2. Variable name and assignment on the same line:

```{r}

iris_long <- iris %>%
  gather(measure, value, -Species) %>%
  arrange(-value)
```

3.Assignment at the end of the pipe with ->:
```{r}

iris %>%
  gather(measure, value, -Species) %>%
  arrange(-value) ->
  iris_long
```

I think this is the most natural to write, but makes reading a little harder: when the name comes first, it can act as a heading to remind you of the purpose of the pipe.


