---
title: "Visualization of spatial information: maps"
author: 
    - "Aitor Ameztegui"
date: "Fall 2021"


output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/udl_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.heigth = 2.5, figure.width=3.5, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
library(tidyverse)
library(sf)
library(rgdal)

fires <- read.csv('data/fires.csv')
grid <- read_sf('data/Fires_grid_CCAA.shp')
```

# Accessing spatial data

The `sf` package will allow us, among other things, to access information contained in spatial vector layers (points, lines and polygons).

The coolest thing about `sf` it's not that it translates vector layers into `data.frames`. The best thing above all is that **it is fully compatible with `dplyr`**.

```{r, eval = F}
#install.packages('sf')
library(raster)
grid <- read_sf('data/Fires_grid_CCAA.shp')
```
---
# The elements of a map

- Legend: symbology key to understand the elements and data in a map (or plot)
- Scale: representation of the unit ratio map-reality
- North: indication of the geographical norht position to understand the arrangment of the map.

---

# Mapping vector data

```{r, eval = F}
#First we summarise number of fires and burned area by CODCUAD
fire.features.grid <- fires %>%
  filter(YEAR>1994,BAREA>1) %>%
  group_by(CODCUAD,.drop = F) %>%
  summarise(N = n(),BA = sum(BAREA))

#Then we join the fire dataset to the grid
grid %>%
    left_join(fire.features.grid,by=c('Cod_Cuad'='CODCUAD')) %>%
    ggplot() +
        geom_sf(aes(fill=log(BA)))
    
```

---

# Mapping vector data


```{r, echo = F}
#First we summarise number of fires and burned area by CODCUAD
fire.features.grid <- fires %>%
  filter(YEAR>1994,BAREA>1) %>%
  group_by(CODCUAD,.drop = F) %>%
  summarise(N = n(),BA = sum(BAREA))

#Then we join the fire dataset to the grid
grid %>%
    left_join(fire.features.grid,by=c('Cod_Cuad'='CODCUAD')) %>%
    ggplot() +
        geom_sf(aes(fill=log(BA)))
    
```


---
# Reading raster layers

The information in raster format is basically a 2-dimensional matrix, in which each of the grids or pixels is given a numerical value that represents either a category o parameter.

The `raster` package provides utilities for manipulating geographic information in raster format.

```{r eval=FALSE}
# install.packages('raster', dep = TRUE)
library(raster)
dem <- raster('data/dem_rioja.tif')
```

---

# Mapping raster layers

```{r, message = F, eval = F}
library(raster)
dem <- raster('data/dem_rioja.tif')

raster::as.data.frame(dem, xy = TRUE) %>%
    ggplot() +
        geom_raster(aes(x=x,y=y,fill=dem_rioja))

```

---

# Mapping raster layers

```{r, message = F, echo = F}
library(raster)
dem <- raster('data/dem_rioja.tif')

raster::as.data.frame(dem, xy = TRUE) %>%
    ggplot() +
        geom_raster(aes(x=x,y=y,fill=dem_rioja))

```

---

```{r, warning = F, message = F, eval = F}
library(rasterVis)
gplot(dem) +
    geom_tile(aes(fill=value)) 
```

---

# Adding scale and norht arrows

```{r, warning = F, message = F, eval = F}
library(ggspatial)
crs(dem)<-"+proj=utm +zone=30 +ellps=intl +units=m +no_defs"
gplot(dem) +
    geom_tile(aes(fill=value)) +
    annotation_scale(location = "bl", width_hint = 0.2) +         #scale
    annotation_north_arrow(location = "bl", which_north = "true", #north
                         pad_x = unit(0.25, "in"), 
                         pad_y = unit(0.25, "in"),
                         style = north_arrow_fancy_orienteering)
```
