---
title: "Homework Unit 6 - Directory of visualizations"
# date: "Due: December 22"
output: 
  html_document: 
    css: hw.css
    theme: yeti
    toc: true
    toc_float: true
    fig_caption: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
knitr::opts_chunk$set(echo = F, eval = F)
```

## 

[TidyTuesday](https://github.com/rfordatascience/tidytuesday#readme) is a weekly data project aimed at the **R ecosystem**. As this
project was borne out of the **R4DS Online Learning Community**, an emphasis
was placed on understanding how to summarize and arrange data to make
meaningful charts with `ggplot2`, `tidyr`, `dplyr`, and other tools in
the `tidyverse` ecosystem.

Every week they post a *raw dataset*, a chart or article related to that
dataset, and ask the community to explore the data. The intent of Tidy Tuesday is to provide a **safe and supportive forum** for
individuals to practice their wrangling and data visualization skills, get
feedback, explore other's work, and connect with the greater `#RStats`
community! As such I encourage you to join the \`TidyTuesday' community and participate in
one of the events, everyone of all skills is welcome!

For today's practice, we are going to use **one of the
datasets** released by this initiative, and we are going to use the skills
learned in Unit 6 to explore the data **through visualizations.**

## Instructions

For this homework you will use the dataset called
[`World freedom index`](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-02-22/readme.md),
which was published the 22/2/22 and contains data on the political
rights and civil liberties of 105 countries and 15 disputed territories.
The data comes from [**Freedom in the
world**](https://freedomhouse.org/reports/publication-archives)**,** a
publication of the [Freedom House](https://freedomhouse.org/)
organization, which has been [published annually since
1972](https://freedomhouse.org/reports/publication-archives), and is the
standard-setting comparative assessment of global political rights and
civil liberties. You can download the data using this code:

```{r echo = T}

library(tidyverse)

freedom <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-02-22/freedom.csv')

```

You need to answer the following questions using charts built with
`ggplot`based on the examples we have seen in Unit 6. There will be
several ways of visualizing the data to help answer the questions, so
there will be no unique solution. I will value the suitability of the
chart to get insight on the question, the quality of the chart, and the
originality. Remember that you can check any source of data and code,
but you are not allowed to share code with your colleagues (but you can
ask them in the forum if you have a specific problem). You can find a
thorough description of the data and the variables in the [TidyTuesday
website](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-02-22/readme.md#data-dictionary).

## Questions

1.  Which continent had the greatest change in civil liberties (CL) and
    political rights (PR) through time? Is it the trend the same for
    both variables?

    *Hint: since the data are at the country level, you will first need
    to summarise it at the continent level\
    Please note that lower values of `CL`* *and* `PR` *mean higher
    standards of liberties and freedom.*

    ```{r}
    freedom %>% 
      group_by(Region_Name, year)%>% 
      summarise(PR = mean(PR)) %>% 
      ggplot() + 
      geom_line(aes(x= year,y = PR, color = Region_Name))

    freedom %>% 
      group_by(Region_Name, year)%>% 
      summarise(CL = mean(CL)) %>% 
      ggplot() + 
      geom_line(aes(x= year,y = CL, color = Region_Name))
    ```

2.  There is a variable called `Status` which determines if a country is
    Free `F`, Not Free `NF`, or Partially Free `PF`. Make a chart to
    show how the number of free countries has varied in the world
    since 1995. \

    ```{r}
    freedom %>%
      filter(Status == "F") %>%
      group_by(year, Region_Name) %>%
      summarise(n_free=n()) %>%
      ggplot() +
      geom_col(aes(x= year, y = n_free, fill = Region_Name))
    ```

3.  Maybe the total number of free countries is not informative, because
    some new countries have been created since 1995. We can get better
    insights if we plot the **proportion** of free, not free and partial
    free countries for **each year**.

    ```{r}
    freedom %>%
      # filter(Region_Name == "Europe") %>%
      group_by(year, Status) %>%
      summarise(n= n()) %>%
      mutate(n_prop = n/sum(n)) %>%
      ggplot() +
        geom_bar(aes(x= year, y = n_prop, fill = Status),
                 stat = "identity")
    ```

4.  Probably, by analyzing all the world at once, we are missing the
    picture. Let's analyze the data by continent, seeing how the values
    of `CL` and `PR` are distributed across each of the continents.
    Which continent has lower values? And which one has more dispersion
    (i.e. more variability within the continent?

    ```{r}
    freedom %>% 
      group_by(Region_Name, country) %>% 
      summarise(PR = mean(PR)) %>% 
      ggplot()+ 
      geom_boxplot(aes(x=Region_Name, y = PR, group = Region_Name, fill = Region_Name))


    freedom %>% group_by(Region_Name, country) %>% summarise(PR = mean(PR)) %>% ggplot()+ ggridges::geom_density_ridges(aes(x=PR,y=Region_Name,fill=factor(Region_Name)))
    ```

5.  Last, analyze if there is any relationship between the scores
    obtained in CL and in PR, and asses if this relationship depends on
    whether the country is a least developed country or not (variable
    `is_ldc` )

    ```{r}
    freedom %>% 
      group_by(is_ldc, country) %>%
      summarise(PR = mean(PR), CL = mean(CL)) %>% 
      ggplot() + 
      geom_point(aes(x = PR, y = CL, color = factor(is_ldc))) +
      geom_smooth(aes(x = PR, y = CL, color = factor(is_ldc), fill = factor(is_ldc)))
    
    
    
    freedom %>%  
        ggplot() + 
        geom_point(aes(x = CL, y = PR, color = factor(is_ldc), 
                       shape = factor(is_ldc)), size = 1.5) + 
        geom_smooth(aes(x = CL, y = PR, color = factor(is_ldc)),method="loess") 
    ```

# 
