---
title: "Saving plots into image files"
date: "111022 | Data management and visualization with R"
author: 
- "Aitor Ameztegui"
- "Marcos Rodrigues"
bibliography: "references/biblio.bib"
output:
  rmdformats::readthedown:
  use_bookdown: true
  highlight: pygments
number_sections: true
css: lab.css
# runtime: shiny_prerendered
---
  
  
```{r knitr_init, echo=FALSE, cache=FALSE, warning=FALSE}

library(knitr)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


```{r include=FALSE}
library(tidyverse)

options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE     # for regression output
)
fires <- read.csv('data/fires.csv')
```

# Saving single plots

We finally have the necessary tools and knwoledge to create nice visualizations. However, we need to be able to export them into image files. Otherwise we cannot embed them into any document.

When saving a plot to use in another program, you have two basic choices of  output: raster or vector: 

* Vector graphics describe a plot as sequence of operations: draw a line from $(x_1, y_1)$ to $(x_2, y_2)$, draw a circle at $(x_3, x_4)$ with radius $r$. This means that they are effectively 'infinitely' zoomable; there is no loss of detail. The most useful vector graphic formats are pdf and svg.
  
* Raster graphics are stored as an array of pixel colours and have a fixed optimal viewing size. The most useful raster graphic format is png.

Unless there is a compelling reason not to, use vector graphics: they look better in more places. There are two main reasons to use raster graphics:

* You have a plot (e.g. a scatterplot) with thousands of graphical objects (i.e. points). A vector version will be large and slow to render.

* You want to embed the graphic in MS Office. MS has poor support for vector graphics (except for their own DrawingXML format which is not currently easy to make from R), so raster graphics are easier.
  
There are two ways to save output from ggplot2. You can use the standard R approach where you open a graphics device, generate the plot, then close the device: \indexf{pdf}

```R
pdf("output.pdf", width = 6, height = 6)
ggplot(mpg, aes(displ, cty)) + geom_point()
dev.off()
```

This works for all packages, but is verbose. ggplot2 provides a convenient shorthand with `ggsave()`:

```R
ggplot(mpg, aes(displ, cty)) + geom_point()
ggsave("output.pdf")
```

`ggsave()` is optimised for interactive use: you can use it after you've drawn a plot. It has the following important arguments: 

* The first argument, `path`, specifies the path where the image should be 
  saved. The file extension will be used to automatically select the correct 
  graphics device. `ggsave()` can produce `.eps`, `.pdf`, `.svg`, `.wmf`,
  `.png`, `.jpg`, `.bmp`, and `.tiff`.

* `width` and `height` control the output size, specified in inches. If left 
  blank, they'll use the size of the on-screen graphics device.
  
* For raster graphics (i.e. `.png`, `.jpg`), the `dpi` argument controls the resolution of the plot.  It defaults to 300, which is appropriate for most printers, but you may want to use 600 for particularly high-resolution output, 
  or 96 for on-screen (e.g., web) display.

See `?ggsave` for more details.

# Arrange multiple plots

It might be the case we want to merge together several plots into a single plot object or image file. There are several packages able to do so in `ggplot`. Some of the most well-known are `ggarrange` or `cowplot`. In this course we are going to focus on `patchwork`, which, to our understanding, is one the most straighforward packages.


# `patchwork `

The goal of `patchwork` is to make it ridiculously simple to combine separate ggplots into the same graphic. As such it tries to solve the same problem as `gridExtra::grid.arrange()` and `cowplot::plot_grid` but using an API that incites exploration and iteration, and scales to arbitrily complex layouts.

## Basic example

The usage of `patchwork` is simple: just add plots together!

```{r example}
library(ggplot2)
library(patchwork)
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p1 + p2
```

patchwork provides rich support for arbitrarily complex layouts with full alignment. As an example, check out this very readable code for nesting three plots on top of a third:

```{r, message=FALSE}
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))
(p1 | p2 | p3) /
      p4
```

## Learn more
patchwork can do so much more. Check out the guides for learning everything there is to know about all the different features:

- [Getting Started](https://patchwork.data-imaginist.com/articles/patchwork.html)
- [Assembling Plots](https://patchwork.data-imaginist.com/articles/guides/assembly.html)
- [Defining Layouts](https://patchwork.data-imaginist.com/articles/guides/layout.html)
- [Adding Annotation](https://patchwork.data-imaginist.com/articles/guides/annotation.html)
- [Aligning across pages](https://patchwork.data-imaginist.com/articles/guides/multipage.html)