---
title: "EC 01 - Exploratory analysis of forest fires in Spain"
date: "Due: March 7 in class"
output: 
  html_document: 
    css: hw.css
    theme: yeti
    toc: true
    toc_float: true
    fig_caption: false
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
#knitr::opts_chunk$set(eval = FALSE)

```

## Exercise 1

In this exercise you will conduct a *guided* exploratory analysis of forest fire data in Spain. We will use the dataset `fires` which compiles fire records in the period 1974-2015. The dataset contains the following information:

- `CODCUAD`: Grid code. Associates fire occurrence to its spatial location.
- `IDPIF`: Fire identifire. Unique code to associate each fire event with the remaining informatin in the original database.
- `YEAR`: Year of occurrence.
- `MONTH`: Month of occurrence. Integer values from 1 to 12.
- `CAUSE`: Ignition source; 1-Natural; 2-Accident; 3-Negligence; 4-Arson; 5-Unknown; and 6-Restarted
- `CCAA`: Autonomous region where the fire started.

```{r, echo=F}
fires <- as_tibble(read.csv('data/fires.csv'))
fires
```

## Instructions

The exercise consists of a set of 10 questions, valued in 1 point apiece. Each of them must be answered, also including the code employed to address them. To do so, we expect you to deliver an RMarkdown document showing both the answer and the code. Create a project in RStudio cloud with your *name* plus *surname* and *Ex_1*.

## Question list

1. Calculate yearly totals of fire counts and burned area. Disregard CAUSE 6 (restared fires).

2. Make a copy of fires dropping fires in 1900.

3. Which MONTH endured the largest amount of fires? 

4. And burned area?

5. In which MONTH do fires grew larger (percentile 97th)?

6. What CAUSE has the highest average fire size?

7. Which was the largest fire occurred in each CCAA?

8. Which CCAA has the highest amount of natural fires?

9. Which CCAA has the highest proportion of natural fires?

10. Create a new dataframe with the number of fires per month and cell (CODCUAD). Put data of each month into an individual column.



